\chapter{Acknowledgements}
\label{cha:acknowledgements}

There is a reason why acknowledgement sections start with thanks to one's supervisor. When I first came to Tromsø, I was a bumbling aspiring phonologist with a knack for producing cryptic, highly dense texts with a philological bent and little systematic knowledge of either theory or analysis. Reading over the text one last time, I notice that the density and the philological asides have not quite been expunged, but Bruce Morén-Duolljá deserves credit (and my gratitude) for all his contribution to making this thesis a reality. It is thanks to his endless patience and relentless insistence on spelling out and unpacking all the assumptions and consequences of the analysis that my work has approached the realm of the intelligible. Our supervision sessions, lasting up to eight hours, rank among the most intellectually stimulating experiences I have ever had. Although I sometimes approached them with trepidation, I would invariably feel I have become a better phonologist at the end, and Bruce is certainly blameless for whatever leaps of faith and hidden assumptions still remain in this thesis.

Going to Tromsø to do a PhD in phonology was one of the best decisions I could ever make. It was first suggested to me by Patrik Bye, who also made sure I could hook up with the rest of the gang in Rhodes, and he deserves special thanks for it. I am grateful to all the CASTL phonologists over the years. Curt Rice's ethos of finding the right questions to ask and laying out the answers clearly and concisely is something I have tried to emulate (with perhaps less success than I would have wished, especially when it comes to the conciseness). Martin Krämer deserves my thanks for his encouragement during the darkest days of graduate student despair and for his pointed advice. I have tried to keep the page count down\dash you and I both know it could have been much worse. Patrik Bye deserves my gratitude for the many conversations we've had about Celtic phonology and about life in general. Ove Lorentz has always been willing to lend an ear to my half\hyp clueless rants about North Germanic languages, coming up with many valuable suggestions. The brief duration of the overlap of my stay in Tromsø with Christian Uffmann's belies his influence on my way of approaching many problems in phonology. Olga Vaysman has also provided numerous nuggets of advice, both phonological and otherwise. Of course, special thanks must go to fellow phonology students: Sylvia Blaho, Peter Jurgec, Islam Youssef, Dragana Šurkalović, Helene Nordgård Andreassen, Violeta Martínez-Paricio, and Olga Urek. It has always been fun, not to mention useful, to bounce my ideas off some of you guys. Speaking of bouncing off ideas, Katalin Balogné Bérces, David Odden, Péter Rácz, Evgeny Shaulskiy, Márton Sóskuthy, Patrycja Strycharczuk, and Barbara Vogt also deserve my gratitude for the many conversations we've had.

Of course, CASTL is not only phonology. Curt Rice and Marit Westergaard have done an enormous amount of work to make sure everything runs so smoothly. Marit deserves special thanks for going out of her way to support me in the crucial late stages of thesis\hyp writing. Thanks to Peter Svenonius for always being willing to talk me through my questions (he cannot be held responsible for anything stupid that I say about morphosyntax). Thanks also to Gillian Ramchand for valuable linguistic and extralinguistic guidance. For the many hours of banter and mutual support (linguistic and otherwise), thanks are due to Monika Bader, Kristine Bentzen, Pavel Caha, Éva Dékány, Antonio Fábregas, Madeleine Halmøy, Björn Lundquist, Andrea Márkus, Rosmin Mathew, Lucie Medová, Sergey Minor, Tom McFadden, Roksolana Mykhaylyk, Marina Pantcheva, Alexander Pfaff, Yulia Rodina, Alexandra Simonenko, Michal Starke, Sandhya Sundaresan, Kaori Takamine, Tarald Taraldsen, Inna Tolskaya, Øystein Vangsnes, Marleen van de Vate, Anna Wolleb, and Naoyuki Yamato. Finally, Christin Kristoffersen, Gry Gaard, Tore Bentz, Kaori Takamine, and Minjeong Son deserve my gratitude for making the administrative part of my CASTL existence hitch\hyp free.

Apart from CASTL, I found what was almost a second home in the Russian group at the Department of Language and Linguistics. Tore Nesset insisted on meeting me basically the moment I came to Tromsø, and I have never regretted it. Thanks to Tore and Laura Janda for the innumerable ways in which they have helped me (including getting valuable teaching hours under my belt) and for their hospitality. Thanks also to Olga Lyashevskaya, Julia Kuznetsova, Svetlana Sokolova, Anastasia Makarova, and Anna Baydimirova.

Øystein Vangsnes was my boss for six months, and one could not have wished for a better one\dash thanks for the opportunity. Thanks also to Trond Trosterud for his advice and his willingness to help in too many ways to count. I have shared an office with Svenn-Egil Knutsen Duolljá and Elisabeth Scheller, whom I thank for their patience. Elsewhere at the University of Tromsø, thanks are due to Jorun Nordmo and Elisabeth Eriksen for invaluable help with administrative issues. Magnhild Svenheim, Bente Storvestre, and Helene Nordgård Andreassen at the University Library would stop at nothing to give me access to the most obscure literature\dash this thesis could not have been written without their help. At Akademisk Kvarter, Isak Måseide has done more than his fair share to help me swell my office library.

Special thanks is due to my teachers at the Faculty of Philology back at Moscow State University, both in the Department of Theoretical and Applied Linguistics and elsewhere. Sergei Tatevosov was my first supervisor, my fieldwork mentor and the person who introduced me to modern phonology. Vladimir Plungian has been invariably supportive of my choices, even though I know he does not always agree with them. Sergei Knyazev's support in phonetic, phonological, and extralinguistic matters has been invaluable. Vladimir Uspensky and Mati Pentus deserve thanks for whatever mathematical prowess I have been able to acquire despite my decidedly non\hyp mathematical upbringing (but are entirely blameless for the blunders that follow). I have been able to hit the ground running after coming to Tromsø thanks to my knowledge of Swedish, for which I must thank Elena Chekalina, Annelie Schlaug, and Robert Leijon.

In Moscow I was also able to develop my interest in Celtic, which has been a deciding factor in where I ended up going. \emph{Diolch yn arbennig i} Elena Parina, who has taught me so much about Welsh. Thanks also to Tatiana Mikhailova, Anna Muradova, and Natalia O'Shea, with whom I have taken classes in the other Celtic languages, as well as Grigory Bondarenko, Alexander Falileyev, and Nina Zhivlova for their help and encouragement.

In the last few years I have had the privilege of sharing a chat at a conference, or a pint at the pub, or an email or Facebook exchange with many exceptional linguists, phonologists and otherwise, from whom I have learned a great deal. There is always a risk of leaving people out, but thanks are due to David Adger, Kristján Árnason, Gwen Awbery, Jill Beckman, Ryan Bennett, Ricardo Bermúdez-Otero, Julian Bradfield, Paul Boersma, Andrew Carnie, Kirsti Koch Christensen, Abby Cohn, Peredur Glyn Davies, Laura Downing, Elan Dresher, Emily Elfner, Þórhallur Eyþórsson, Paula Fikkert, Janet Grijzenhout, Maria Gouskova, Atle Grønn, Aðalsteinn Hákonarson, Daniel Currie Hall, Silke Hamann, Mike Hammond, S.~J. Hannahs, Sam Hellmuth, Ben Hermans, Steve Hewitt, Patrick Honeybone, Janne Bondi Johannessen, Junko Itô, Dan Karvonen, Darya Kavitskaya, Yuni Kim, John Kingston, Paul Kiparsky, Alexei Kochetov, Björn Köhnlein, Terje Lohndal, Jim McCloskey, Fiona McLaughlin, Armin Mester, Andrew Nevins, Máire Ní Chiosáin, David Odden, Marc van Oostendorp, Jaye Padgett, Markus Pöchtrager, Péter Rácz, Keren Rice, Michael Rießler, Catherine Ringen, Nathan Sanders, Tobias Scheer, Ranjan Sen, Michael Schäfer, Koen Sebregts, Norval Smith, Márton Sóskuthy, Patrycja Strycharczuk, Sally Thomason, Jochen Trommer, Ruben van de Vijver, David Willis, and up to seven anonymous reviewers.

I thank my friends in Tromsø, Moscow, Oxford, and elsewhere for staying in touch and providing much\hyp needed support. Thanks to Natalia Aralova, Maria Artamonova, Juliana Dresvina, David Erschler, Alexandra Faynveyts, Dmitri Hrapof, Olga Lyashevskaya, Tom McFadden and Sandhya Sundaresan, Elena Parina, Dmitri Sitchinava, Evgeny Smirnov, Svetlana Sokolova, Inna Tolskaya and Dmitri Shcherbin, and Alexander Zhulin and Maria Lavrenchenko.

Finally, I owe a huge debt of gratitude to my family for their support, encouragement, and most of all patience. My mother Nadezhda was always there when I needed her. My brother Alexander not only supported me emotionally but also often helped me out with access to literature while I was an autodidact. I have always felt the support of my father Vladimir.

My wife Irina followed me to Tromsø (and further) and supported me all the way despite my tendency to bury myself in work and become all too irritable all too often. Both Irina and my daughter Dina showed angelic patience with me not really being there for a long time, and I apologize to them both. I dedicate this thesis to them.